lifespan_eda

### 2024-03-20 eda/lifespan_eda.Rmd
parsing through results, fix na.omit

Visualizing euler distributions
```{r}
cn.df.unique %>%
  ggplot() +
  geom_boxplot(aes(x=study, y=surfholes))
```

@@

```{r}
cn.df.eulerfilt <- cn.df.unique %>%
  dplyr::filter(!is.na(surfholes))

@@

# left_join(cn.df.eulerfilt, study_mad) %>%
#   ggplot() +
#   geom_boxplot(aes(x=study, y=surfholes)) +
#   geom_point(aes(x=study, y=(study_euler_median+2*study_euler_madconst1)), color="red")

left_join(cn.df.eulerfilt, study_mad) %>%
  mutate(scan_pass = ifelse(surfholes <= euler_threshold, "yes", "no")) %>%
  group_by(study) %>%
  summarise(fail_percent = (sum(scan_pass == "no") / n())*100) %>%
  summarise(max_fail = max(fail_percent),
            min_fail=min(fail_percent),
            mean_fail=mean(fail_percent))

```

@@

```{r}
cn.df.site <- cn.df.eulerfilt2 %>% 
  dplyr::filter(site != "NA") %>%
  mutate(study = droplevels(study),
         site = gsub(" ", "_", trimws(site))) %>%
  unite(site, study, site, sep = "_", remove = FALSE) %>% #so sites with the same name don't get combined across studies!
  na.omit()
n_unique(cn.df.site$participant)
n_unique(cn.df.site$participant) == nrow(cn.df.site)

@@

```{r}
ggplot(cn.df.site) +
  geom_bar(aes(x=site, fill=sex))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

@@

```{r}
fwrite(cn.df.site, file=paste0(base_path, "/data/lifespan_CN_site-level_euler.csv"))
```
write out imagen data for testing
```{r}
imagen.df <- cn.df.site %>%
  dplyr::filter(study=="IMAGEN")
# skim(imagen.df)

fwrite(imagen.df, file=paste0(base_path, "/data/imagen_CN.csv"))
```

troubleshooting combat

```{r, eval=FALSE}
gamlss(formula = rh_Vol_posteriorcingulate ~ pb(age_days) + sexMale + sex.age + site, sigma.formula= ~ pb(age_days) + sexMale + site, data=imagen.df)
```

looking into whether all global measures should be log-transformed to account for skew (getting neg vals for ventricles with combat-GAM)

```{r}
global_vols <- cn.df.site %>%
  dplyr::select(sex, log_age, all_of(vol_list_global)) %>%
  pivot_longer(cols=c(-sex, -log_age), names_to = "pheno", values_to = "volume") %>%
  mutate(log_vol = log(volume, base=10)) #test log-transform

ggplot(global_vols) +
  geom_point(aes(x=log_age, y=volume, color=sex), alpha=0.4) +
  facet_wrap(~pheno, scales="free_y")

ggplot(global_vols) +
  geom_histogram(aes(x=volume, fill=pheno), alpha=0.5) +
  facet_wrap(~pheno, scales="free")

ggplot(global_vols) +
  geom_histogram(aes(x=log_vol, fill=pheno), alpha=0.5) +
  facet_wrap(~pheno, scales="free")
```

@@

nrow(cn.df.imp1) - nrow(cn.df.imp)
# unique(cn.df.imp$site)
# table(cn.df.imp$site, cn.df.imp$study)
# sum(is.na(cn.df.imp$site))
```

@@

```{r}
ggplot(cn.df.imp) +
  geom_bar(aes(x=site, fill=sex))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

@@

## Visualize Harmonized Data

comparing ventricle vals across harmonization pipelines (neg vals in `lifespan-CN-imp-sites_cf.gam_data.csv`), may need to log-transform
```{r}
lifespan_imp_gam.df <- fread(paste0(base_path, "/data/lifespan/lifespan-CN-imp-sites_cf.gam_data.csv"), stringsAsFactors = TRUE)
lifespan_imp_gamlss.df <- fread(paste0(base_path, "/data/lifespan/lifespan-CN-imp-sites_cf.gamlss_data.csv"), stringsAsFactors = TRUE)

lifespan_gam.df <- fread(paste0(base_path, "/data/lifespan/lifespan-CN-site-level_cf.gam_data.csv"), stringsAsFactors = TRUE)
lifespan_gamlss.df <- fread(paste0(base_path, "/data/lifespan/lifespan-CN-site-level_cf.gamlss_data.csv"), stringsAsFactors = TRUE)
```

```{r}
#mark negative vals
# cn.df.site <- as.data.frame(cn.df.site)
neg1 <- cn.df.site %>%
  dplyr::filter(Ventricles <= 0)
ven_raw <- ggplot() +
  geom_point(data=cn.df.site, aes(x=log_age, y=Ventricles, color=sex), alpha =0.3) +
  geom_point(data=neg1, aes(x = log_age, y=Ventricles), color='red', alpha=0.6) + 
  ggtitle("Raw Data")

neg2 <- lifespan_gam.df[lifespan_gam.df$Ventricles <=0]
ven_gam <- ggplot() +
  geom_point(data=lifespan_gam.df, aes(x=log_age, y=Ventricles, color=sex), alpha =0.3) +
  geom_point(data=neg2, aes(x = log_age, y=Ventricles), color='red', alpha=0.6) + 
  ggtitle("ComBat-GAM")

neg3 <- lifespan_gamlss.df[lifespan_gamlss.df$Ventricles <=0]
ven_gamlss <- ggplot(lifespan_gamlss.df) +
  geom_point(aes(x=log_age, y=Ventricles, color=sex), alpha =0.3) +
  geom_point(data=neg3, aes(x = log_age, y=Ventricles), color='red', alpha=0.6) + 
  ggtitle("ComBatLS")

# cn.df.imp <- as.data.frame(cn.df.imp)
neg4 <- cn.df.imp %>%
  dplyr::filter(Ventricles <= 0)
ven_raw.imp <- ggplot(cn.df.imp) +
  geom_point(aes(x=log_age, y=Ventricles, color=sex), alpha =0.3) +
  geom_point(data=neg4, aes(x = log_age, y=Ventricles), color='red', alpha=0.6) + 
  ggtitle("Raw Data - All Studies")

neg5 <- lifespan_imp_gam.df[lifespan_imp_gam.df$Ventricles <=0]
ven_gam.imp <- ggplot(lifespan_imp_gam.df) +
  geom_point(aes(x=log_age, y=Ventricles, color=sex), alpha =0.3) +
  geom_point(data=neg5, aes(x = log_age, y=Ventricles), color='red', alpha=0.6) + 
  ggtitle("ComBat-GAM - All Studies")

neg6 <- lifespan_imp_gamlss.df[lifespan_imp_gamlss.df$Ventricles <=0]
ven_gamlss.imp <- ggplot(lifespan_imp_gamlss.df) +
  geom_point(aes(x=log_age, y=Ventricles, color=sex), alpha =0.3) +
  geom_point(data=neg6, aes(x = log_age, y=Ventricles), color='red', alpha=0.6) + 
  ggtitle("ComBatLS - All Studies")

neg_ventricle_plots <- ggarrange(ven_raw, ven_gam, ven_gamlss, ven_raw.imp, ven_gam.imp, ven_gamlss.imp, common.legend = TRUE)
#ggsave(paste0(base_path, "figs/lifespan_ventricle_neg.jpg"), plot=neg_ventricle_plots)
```

## Combat Troubleshooting
having some issues with combat on cubic, trying locally.

>>>>

Remove infants (anyone younger than 3 yrs)

```{r}
cn.df.noinf <- cn.df.unique %>%
  dplyr::filter(age_days >= (365.25*3)) %>%
  mutate(study = droplevels(study),
         site = droplevels(site))
nrow(cn.df.unique) - nrow(cn.df.noinf)
n_unique(cn.df.unique$study) - n_unique(cn.df.noinf$study)
min(cn.df.noinf$age_days)
```

Visualizing euler distributions
```{r}
cn.df.noinf %>%
  ggplot() +
  geom_boxplot(aes(x=study, y=surfholes))
```

@@

```{r}
cn.df.eulerfilt <- cn.df.noinf %>%
  dplyr::filter(!is.na(surfholes))

@@

left_join(cn.df.eulerfilt, study_mad) %>%
  mutate(scan_pass = ifelse(surfholes <= euler_threshold, "yes", "no")) %>%
  group_by(study) %>%
  summarise(fail_percent = (sum(scan_pass == "no") / n())*100) %>%
  summarise(max_fail = max(fail_percent),
            min_fail=min(fail_percent),
            mean_fail=mean(fail_percent))
```

@@

```{r}
cn.df.site <- cn.df.eulerfilt2 %>% 
  dplyr::filter(site != "NA") %>%
  mutate(study = droplevels(study),
         site = gsub(" ", "_", trimws(site))) %>%
  unite(site, study, site, sep = "_", remove = FALSE) %>% #so sites with the same name don't get combined across studies!
  dplyr::select(!c(INDEX.ID, country)) %>%#drop index ID for filtering
  na.omit()

n_unique(cn.df.site$participant)
n_unique(cn.df.site$participant) == nrow(cn.df.site)

@@

@@

```{r}
fwrite(cn.df.site, file=paste0(base_path, "/data/lifespan_CN_site-level_euler.csv"))
```

@@

nrow(cn.df.imp1) - nrow(cn.df.imp)
n_unique(cn.df.imp$site)
n_unique(cn.df.imp$study)
# table(cn.df.imp$site, cn.df.imp$study)
# sum(is.na(cn.df.imp$site))
```

@@

@@
